Dynamics of Laterally Propagating Flames in X-ray Bursts

### 2019-10-04 xrb/paper.tex
added more marginpar's

\author{Yuri Cavecchi}
\affiliation{Princeton University}

@@

Since the neutron star rotates, we work in a corotating frame, with
angular velocity $\Omegab$.  Further, for the two-dimensional
simulations presented here, we work in axisymmetric coordinates,
but we advect a third component of velocity, coming out of the
simulation plane, that participates in the Coriolis force (sometimes
described as a 2.5D simulation).  We will take the
\castro\ $x$-coordinate to be the cylindrical radial coordinate with
corresponding velocity $u$, the \castro\ $y$-coordinate to be the
cylindrical vertical coordinate with corresponding velocity $v$, and the
\castro\ $z$-coordinate to be the cylindrical azimuthal coordinate,
with corresponding velocity $w$.  A righthanded coordinate system has
positive $w$ pointing out of the page.  We will take $\Omegab =
\Omega_0 \hat{\bf y}$ for the angular rotation rate, and $\gb = -g
\hat{\bf y}$ for the gravitational acceleration, with $g$ constant.
With these choices, the Coriolis force is:
\begin{equation}
-2\rho \Omegab \times \Ub =
   -2\rho \left ( \Omega_0 w \hat{x} - \Omega_0 u \hat{z} \right )
\end{equation}
We will neglect the centrifugal force---with our plane-parallel
geometry, this will act only in the lateral direction, and is not
expected to greatly affect the dynamics.  Carrying the Coriolis force
allows us to capture the geostrophic balance\MarginPar{ref} that sets
up via lateral hydrostatic equilibrium.

@@

All simulations use adaptive mesh refinement to refine on the
atmosphere (leaving the space between the top of the atmosphere and
upper boundary at low resolution).  In addition to the base grid, we
use up to 3 refinement levels, the first one a factor of 4 finer than
the previous and the remaining each a factor of 2 finer than the
previous.  \castro\ subcycles in time, so the finer grids are evolved
with a finer timestep than the coarse grids.  Occasionally the
timestep chosen at the start of a cycle will violate the CFL condition
during the advance of the finer grids.  In this case, we restart the
finer grid evolution with a smaller timestep, subcycling within the
larger timestep hierarchy.  We use a CFL number of 0.8 for our
simulations.  \MarginPar{true for all?} The base grid for our standard
simulations is $768\times 192$ zones and the corresponding finest grid
when 3 refinement levels are added would be $12288\times 3072$ zones.
Our standard domain is $1.2288\times 10^5~\mathrm{cm} \times
3.072\times 10^4~\mathrm{cm}$, so we have $10~\mathrm{cm}$ resolution
on the finest grid.  We only refine the fuel layer in the left half of
the domain at the highest resolution (and only down to densities of
$2.5\times 10^4~\gcc$), since this is where we expect the flame to
propagate.  At the start of the simulation, $3.4\%$ of the domain is
at the finest resolution.  This increases to $7.4\%$ by the end of the
simulation, because of the increase in the scale height of the atmosphere
behind the flame.

@@

The speed and thickness of a laminar helium flame is determined by the
energy generation rate and conductivity and scale roughly as \MarginPar{need a ref}
\begin{equation}
v_f \approx \sqrt{k_\mathrm{th} \dot{\epsilon}}
\end{equation}
and the flame width scales as:
\begin{equation}
\lambda_f = \sqrt{\frac{k_\mathrm{th}}{\dot{\epsilon}}}
\end{equation}
At the densities we consider in this simulation, the pure He flame
speed is quite slow and would require a prohibitively long integration
time to see significant evolution of the burning.  We therefore
consider boosted flames in this first paper, to accelerate the burning
and allow us to understand the qualitative effects of laterally
propagating flames.  To boost the flame, while keeping the thickness
the same, we can multiply both the burning rate and the conductivity
by the same factor.  For our standard calculations, we choose 10 for
each, to give a flame 10$\times$ faster.  We call this the ``10/10''
flame.  We will also do a simulation with the reactions and
conductivity both boosted by 5, the ``5/5'' flame.

To understand the time and length scales involved in flame
propagation, we do a 1D simulation of a laminar flame using our
microphysics.  Figure~\ref{fig:flame} shows the flame properties for a \AssignTo{Kiran and Mike will remake with the 10/10 boosted flame}
He flame using the standard conductivities and our simple reaction
network.  This flame had a density of $2\times 10^6~\gcc$ and \MarginPar{double check}
temperature of $5\times 10^7$~K.  We note that this flame speed is
about $2\times 10^4~\cms$, the flame width is about 20~cm, and it
takes about 0.1~s to settle into a sustained flame.  Note this speed
is quite small compared to the speed estimated in
\citet{spitkovsky2002}.  We expected a multidimensional flame to
accelerate due to hydrodynamics interactions (wrinkling, turbulence
interactions, directed flows feeding fuel into the flame, ...).

@@

\begin{figure}[t]
\centering
\plotone{model}
\caption{\label{fig:sketch} A schematic of our model setup.}
\end{figure}

We wish to create an initial atmosphere consisting of a hot
``post-flame'' region and a cooler atmosphere that the flame will
laterally propagate into.  We put the hot region at the very left of
the domain (the origin of the axisymmetric coordinates).  To create
these initial conditions, we produce two different hydrostatic models,
a ``hot'' model that will represent the perturbation that drives the
flame and a ``cool'' model that will represent the state ahead of the
flame.  These will have different scale heights.  To create these
models, we break the vertical structure of the atmosphere into 4
layers: the underlying star, a ramp-up to the base of the
accreted atmosphere, a fuel layer representing the bulk of the
atmosphere where the flame will propagate, and an outer, low density,
isothermal buffer above the atmosphere that allows us
to capture expansion and explosive dynamics.  This is sketched in
Figure~\ref{fig:sketch}.

@@

We note that the entire simulation framework for these calculations is
freely available in the \castro\ github repository.

\begin{deluxetable}{lccccc}
\tablecaption{\label{table:sim_names} Simulaiton parameters.}
\tablehead{\colhead{name} & \colhead{reaction boost} & \colhead{conductivity boost} & \colhead{fine grid resolution} & \colhead{rotation rate}  & \colhead{network}}
\startdata
10/10 & 10 & 10 & 10~cm & 2000~Hz & {\tt aprox13} \\
10/10-slow & 10 & 10 & 10~cm & 1000~Hz & {\tt aprox13} \\
10/10-lores & 10 & 10 & 20~cm & 2000~Hz & {\tt aprox13} \\
5/5 & 5 & 5 & 10~cm & 2000~Hz & {\tt aprox13} \\
10/10-iso7 & 10 & 10 & 10~cm & 2000~Hz & {\tt iso7} \\
\enddata
\end{deluxetable}

@@

Figure~\ref{fig:time_series} shows the time

@@

\MarginPar{do we want to do a low res run?}

@@

These studies complement the

@@

There are several changes we will pursue to the algorithm used to
model these XRBs.  First, as shown in \citet{castro_sdc}, we have
developed a fourth-order (in space and time) method for coupling
hydrodynamics and reactions that should greatly improve the accuracy
of the simulations.  We expect that by using this new high-order
algorithm we can drop a level of refinement from the simulations while
still accurately modeling the evolution.  We have also ported
\castro\ to GPUs, giving an order of magnitude performance boost on
nodes with both CPUs and GPUs.  The GPU-enabled algorithm will
increase the performance, allowing us to run bigger domains and/or
eliminate the boosting used here.

>>>>

\author{Yuri Cavecchi}
\affiliation{Department of Astrophysical Sciences, Princeton University, Peyton Hall, Princeton, NJ 08544}
\affiliation{Mathematical Sciences and STAG Research Centre, University of Southampton, SO17 1BJ}

@@

Since the neutron star rotates, we work in a corotating frame, with
angular velocity $\Omegab$.  Further, for the two-dimensional
simulations presented here, we work in axisymmetric coordinates,
but we advect a third component of velocity, coming out of the
simulation plane, that participates in the Coriolis force (sometimes
described as a 2.5D simulation).  We will take the
\castro\ $x$-coordinate to be the cylindrical radial coordinate with
corresponding velocity $u$, the \castro\ $y$-coordinate to be the
cylindrical vertical coordinate with corresponding velocity $v$, and the
\castro\ $z$-coordinate to be the cylindrical azimuthal coordinate,
with corresponding velocity $w$.  A righthanded coordinate system has
positive $w$ pointing out of the page.  We will take $\Omegab =
\Omega_0 \hat{\bf y}$ for the angular rotation rate, and $\gb = -g
\hat{\bf y}$ for the gravitational acceleration, with $g$ constant.
With these choices, the Coriolis force is:
\begin{equation}
-2\rho \Omegab \times \Ub =
   -2\rho \left ( \Omega_0 w \hat{x} - \Omega_0 u \hat{z} \right )
\end{equation}
We will neglect the centrifugal force---with our plane-parallel
geometry, this will act only in the lateral direction, and is not
expected to greatly affect the dynamics.  Carrying the Coriolis force
allows us to capture the geostrophic balance\MarginPar{ref, Spitkovsky 2002?} that sets
up via lateral hydrostatic equilibrium.

@@

All simulations use adaptive mesh refinement to refine on the
atmosphere (leaving the space between the top of the atmosphere and
upper boundary at low resolution).  In addition to the base grid, we
use up to 3 refinement levels, the first one a factor of 4 finer than
the previous and the remaining each a factor of 2 finer than the
previous.  \castro\ subcycles in time, so the finer grids are evolved
with a finer timestep than the coarse grids.  Occasionally the
timestep chosen at the start of a cycle will violate the CFL condition
during the advance of the finer grids.  In this case, we restart the
finer grid evolution with a smaller timestep, subcycling within the
larger timestep hierarchy.  We use a CFL number of 0.8 for our
simulations.  
% \MarginPar{true for all?} 
The base grid for our standard
simulations is $768\times 192$ zones and the corresponding finest grid
when 3 refinement levels are added would be $12288\times 3072$ zones.
Our standard domain is $1.2288\times 10^5~\mathrm{cm} \times
3.072\times 10^4~\mathrm{cm}$, so we have $10~\mathrm{cm}$ resolution
on the finest grid.  We only refine the fuel layer in the left half of
the domain at the highest resolution (and only down to densities of
$2.5\times 10^4~\gcc$), since this is where we expect the flame to
propagate.  At the start of the simulation, $3.4\%$ of the domain is
at the finest resolution.  This increases to $7.4\%$ by the end of the
simulation, because of the increase in the scale height of the atmosphere
behind the flame.

@@

The speed and thickness of a laminar helium flame are determined by the
energy generation rate and conductivity, and scale roughly as \MarginPar{need a ref}
\begin{equation}
v_f \approx \sqrt{k_\mathrm{th} \dot{\epsilon}}
\end{equation}
and the flame width scales as:
\begin{equation}
\lambda_f = \sqrt{\frac{k_\mathrm{th}}{\dot{\epsilon}}}
\end{equation}
At the densities we consider in this simulation, the pure He flame
speed is quite slow and would require a prohibitively long integration
time to see significant evolution of the burning.  We therefore
consider boosted flames in this first paper, to accelerate the burning
and allow us to understand the qualitative effects of laterally
propagating flames.  To boost the flame, while keeping the thickness
the same, we can multiply both the burning rate and the conductivity
by the same factor.  For our standard calculations, we choose 10 for
each, to give a flame 10$\times$ faster.  We call this the ``10/10''
flame.  We will also do a simulation with the reactions and
conductivity both boosted by 5, the ``5/5'' flame.

To understand the time and length scales involved in flame
propagation, we do a 1D simulation of a laminar flame using our
microphysics.  Figure~\ref{fig:flame} shows the flame properties for a 
% \AssignTo{Kiran and Mike will remake with the 10/10 boosted flame}
He flame using the standard conductivities and our simple reaction
network.  This flame had a density of $2\times 10^6~\gcc$ and 
% \MarginPar{double check}
temperature of $5\times 10^7$~K.  We note that this flame speed is
about $2\times 10^4~\cms$, the flame width is about 20~cm, and it
takes about 0.1~s to settle into a sustained flame.  Note this speed
is quite small compared to the speed estimated in
\citet{spitkovsky2002}.  We expected a multidimensional flame to
accelerate due to hydrodynamics interactions (wrinkling, turbulence
interactions, directed flows feeding fuel into the flame, ...).

@@

% \begin{figure}[t]
% \centering
% \plotone{model}
% \caption{\label{fig:sketch} A schematic of our model setup.}
% \end{figure}

We wish to create an initial atmosphere consisting of a hot
``post-flame'' region and a cooler atmosphere that the flame will
laterally propagate into.  We put the hot region at the very left of
the domain (the origin of the axisymmetric coordinates).  To create
these initial conditions, we produce two different hydrostatic models,
a ``hot'' model that will represent the perturbation that drives the
flame and a ``cool'' model that will represent the state ahead of the
flame.  These will have different scale heights.  To create these
models, we break the vertical structure of the atmosphere into 4
layers: the underlying star, a ramp-up to the base of the
accreted atmosphere, a fuel layer representing the bulk of the
atmosphere where the flame will propagate, and an outer, low density,
isothermal buffer above the atmosphere that allows us
to capture expansion and explosive dynamics.  
% This is sketched in
% Figure~\ref{fig:sketch}.

@@

We note that the entire simulation framework for these calculations is
freely available in the \castro\ github repository\footnote{\url{https://github.com/AMReX-Astro/Castro}}.

\begin{deluxetable}{lccccc}
\tablecaption{\label{table:sim_names} Simulation parameters.}
\tablehead{\colhead{name} & \colhead{reaction boost} & \colhead{conductivity boost} & \colhead{fine grid resolution} & \colhead{rotation rate}  & \colhead{network}}
\startdata
10/10 & 10 & 10 & 10~cm & 2000~Hz & {\tt aprox13} \\
10/10-slow & 10 & 10 & 10~cm & 1000~Hz & {\tt aprox13} \\
10/10-lores & 10 & 10 & 20~cm & 2000~Hz & {\tt aprox13} \\
5/5 & 5 & 5 & 10~cm & 2000~Hz & {\tt aprox13} \\
10/10-iso7 & 10 & 10 & 10~cm & 2000~Hz & {\tt iso7} \\
\enddata
\end{deluxetable}

@@

Figure~\ref{fig:time_series} shows the time \MarginPar{this plot needs a description}

@@

\MarginPar{talk about the low res run?}

@@

These studies complement the\MarginPar{???? Finish this sentence}

@@

There are several changes we will pursue to the algorithm used to
model these XRBs.  First, as shown in \citet{castro-sdc}, we have
developed a fourth-order (in space and time) method for coupling
hydrodynamics and reactions that should greatly improve the accuracy
of the simulations.  We expect that by using this new high-order
algorithm we can drop a level of refinement from the simulations while
still accurately modeling the evolution.  We have also ported
\castro\ to GPUs, giving an order of magnitude performance boost on
nodes with both CPUs and GPUs.  The GPU-enabled algorithm will
increase the performance, allowing us to run bigger domains and/or
eliminate the boosting used here.
